A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-28 natural history of scavenging.tex
Andrew's edits

\documentclass[a4paper,12pt]{article}
\usepackage{ecography}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{xfrac}

@@

@@

\begin{abstract} 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. 
These can be thought of as some of the principal parameters in optimal foraging theory namely, encounter rate, handling time and prey availability. 
We use these components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focusing on the dominant vertebrate groups of the land, sea and air. 
The result is a document on the natural history of scavenging, the first to our knowledge. 
Our idea of a scale of scavenging can be applied to any species at any time to judge the importance of this behaviour in its diet. 
\end{abstract}

@@

\subsection*{The Challenges of Scavenging} 
The chief hurdle to scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. 
However, species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and process the, typically decaying, carcass replete with micro-organism derived toxins  \citep{ruxton2014fruit}. 
These characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics but are outside the selection pressures on the scavenger.
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the likely importance of carrion in its diet.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Summary_figure/Summary_figure_Landscape.pdf}
\caption{Factors influencing the proportion of scavenging in a vertebrates' diet. Blue lines indicate a reduction in the factor and red lines indicate an increase.}
%TG: Probably needs some more explanations...
\label{Summary_figure}
\end{figure}

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all increase the encounter rate between a scavenger and the carcasses it is searching for. 
Alternatively, a species can reduce its metabolic requirements so that it can survive long periods between meals. 
We would thus expect selection pressures to act on the various traits that govern these parameters.

@@

\subsection*{Metabolism}
Because of the sporadic nature of carrion we would expect adaptations that reduce energetic costs of maintenance to be selected for in scavengers as it would allow for longer inter-feeding periods. 
Extant reptiles possess an advantage here, in that over the course of a year their food requirements can be 30 times lower than an endotherm of equal size \citep{Nagy1621}.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit exceedingly low maintenance metabolisms, and most can survive on a few scant feedings per year.
It is, therefore, possible for snakes to rely largely on infrequent, less energy-rich meals."
In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodiles and their allies because a sit and wait strategy is viable for an ectotherm \citep{forrest2003evidence,moleon2015carcass}. 
\cite{carey1982temperature} found ectothermic sharks have the ability to exist weeks between meals because they focus on the energy-rich sections of carcasses (see Handling Time). 
Endotherms have also evolved physiological mechanisms that allow them to depress their otherwise high metabolic rates at certain times e.g. vultures who do so while resting at their roost and during periods of food deprivation \citep{bahat1998nocturnal}.

@@

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal  movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, and references therein). 
For instance, \cite{alexander2004bipedal} shows  that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with their argument that long distance endurance running, often used by humans for hunting, was not an important feature of hominin scavenging.
They arrive at this conclusion on the basis of the high cost of running coupled with the fact that dangerous competitors will tend to arrive beforehand.

@@

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent studies have shown that the energetic demands of such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{carbone2011intra,kane2016body}.

@@

Species capable of flight have effectively added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
The flying pterosaurs also convergently evolved large orbits and optic lobes \citep{witton2013pterosaurs}. 
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts through vision \citep{AR:AR22815}.

Existing in this '3D' environment also means being able to gather a wealth of information from other foragers, be they conspecifics or otherwise \citep{jackson2008effect}.
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them to forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.
Information transfer of this kind is typically inadvertent and as a consequence no complex social interactions are required, simply the ability to recognise a successful forager.
Thus, it seems probable that scrounging behaviours were seen in the flying pterosaurs as well.

@@

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
The geological record shows the Earth has undergone radical fluctuations in temperature over time.
This will have had a significant bearing on the availability and persistence of carrion.
To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was on average at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
Moreover, oceanic productivity and habitat structure are all impacted by climactic conditions.
The impacts these can have on scavengers have been empirically supported e.g. \cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".
This is a sobering thought given the impact we humans are having on the Earth's climate.

@@

\section*{Handling Time}
Since the food a scavenger depends upon is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be subject to decay as well as competition.
So being able to overcome competitors and maximise the nutrient gain from the remnants are all essential parts of carcass handling time.

@@

\subsection*{Food Processing}
Aside from coping with competitors, another vital component of carrion handling time is the ability to maximise the energy gain from the remains. 
At whale carcases, white and blue sharks are known to preferentially feed on the blubber layer \citep{long1996sharks}. 
Blubber is an energy rich portion of the carcass that can allow a shark to survive for 1.5 months on 30 kg of the material \citep{carey1982temperature}. 
On land many scavengers utilize late-stage carcass material that is less subject to decomposition and may be unavailable to other competitors, for example bone.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{timeline_figure/timeLine.pdf}
\caption{The diversity of scavengers through time. Each species has either direct evidence for it being a scavenger or would be positioned high up on our scavenging scale.}
\label{Timeline}
\end{figure}

@@

\section*{Acknowledgments}

A lot of people are to thank here.

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{ecography}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{xfrac}
\usepackage{lineno}

@@

\nolinenumbers

@@

\begin{abstract} 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. 
These can be thought of as some of the principal parameters in optimal foraging theory namely, encounter rate, handling time and prey availability. 
We use these components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focusing on the dominant vertebrate groups of the land, sea and air. 
The result is a synthesis on the natural history of scavenging, the first to our knowledge. 
Our idea of a scale of scavenging can be applied to any species at any time to judge the importance of this behaviour in its diet. 
\end{abstract}

@@

\subsection*{The Challenges of Scavenging} 
The chief hurdle to scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. 
However, species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and efficiently process the, typically decaying, carcass replete with micro-organism derived toxins  \citep{ruxton2014fruit}. 
These characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics but are outside the selection pressures on the scavenger.
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the likely importance of carrion in its diet.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Summary_figure/Summary_figure.pdf}
\caption{Factors influencing the proportion of scavenging in a vertebrates' diet. Blue lines indicate a reduction in the factor and red lines indicate an increase.}
%TG: Probably needs some more explanations...
\label{Summary_figure}
\end{figure}

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all determine the encounter rate between a scavenger and the carcasses it is searching for. 
Alternatively, a species can reduce its metabolic requirements so that it can survive long periods between meals. 
We would thus expect selection pressures to act on the various traits that govern these parameters.

@@

\subsection*{Metabolism}
Because of the sporadic nature of carrion we would expect adaptations that reduce energetic costs of maintenance to be selected for in scavengers as it would allow for longer inter-feeding periods. 
Extant reptiles possess an advantage here, in that over the course of a year their food requirements can be 30 times lower than an endotherm of equal size \citep{Nagy1621}.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit exceedingly low maintenance metabolisms, and most can survive on a few scant feedings per year.
It is, therefore, possible for snakes to rely largely on infrequent, less energy-rich meals."
In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodiles and their allies \citep{forrest2003evidence,moleon2015carcass}. 
\cite{carey1982temperature} found ectothermic sharks have the ability to go weeks between meals because they focus on the energy-rich sections of carcasses (see Handling Time). 
Endotherms have also evolved physiological mechanisms that allow them to depress their otherwise high metabolic rates at certain times e.g. vultures who do so while resting at their roost and during periods of food deprivation \citep{bahat1998nocturnal}.

@@

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal  movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, and references therein). 
For instance, \cite{alexander2004bipedal} shows  that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement calculated by allometric scaling relationships.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with their argument that long distance endurance running, often used by humans for hunting, was not an important feature of hominin scavenging.
They arrive at this conclusion on the basis of the high cost of running coupled with the fact that dangerous competitors will tend to arrive beforehand.

@@

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent work has shown that the energetic demands of such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{kane2016body}.

@@

Species capable of flight have added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
In this way they are effectively cheating the 2D system by gaining a bird's eye view which has has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
The flying pterosaurs also convergently evolved large orbits and optic lobes \citep{witton2013pterosaurs}. 
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts through vision \citep{AR:AR22815}.

Existing in this environment also means being able to gather a wealth of information from other foragers, be they conspecifics or otherwise \citep{jackson2008effect}.
These interactions are properly three dimensional in the sense of \cite{pawar2012dimensionality} because both producer and scrounger are in the air \citep{dall2005information}. 
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them to forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.
Information transfer of this kind is typically inadvertent and as a consequence no complex social interactions are required, simply the ability to recognise a successful forager.
Thus, it seems probable that scrounging behaviours were seen in the flying pterosaurs as well.

@@

Perhaps the greatest environmental driver of scavenging tendency is that of temperature. 
The geological record shows the Earth has undergone radical fluctuations in temperature over time.
This will have had a significant bearing on the availability and persistence of carrion.
To illustrate the point, a 10$^{\circ}$C increase in ambient temperature can double carcass decomposition rates \citep{parmenter2009carrion} and geological evidence indicates that the Mesozoic Earth was on average at least 6 $^{\circ}$C warmer than now \citep{sellwood2006mesozoic}.
In terms of specific habitats, it has been shown that decomposition is greater in warm and moist areas versus more xeric ones \citep{beasley2015vertebrates}.
Moreover, oceanic productivity and habitat structure are all impacted by climactic conditions.
The impacts these can have on scavengers have been empirically supported e.g. \cite{beasley2015vertebrates} who point to a series of studies showing how microbes and invertebrates benefit at higher temperatures to the detriment of vertebrate scavengers such that ``above 20$^{\circ}$C vertebrates were able to detect and consume only 19\% of small-mammal carcasses, whereas at temperatures below 18$^{\circ}$C, vertebrates consumed 49\% of carcasses".

@@

\section*{Handling Time}
Since the food a scavenger depends upon is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing owing to the activity of predators and other scavengers, or, if present, will be subject to decay as well as competition.
So being able to overcome competitors and maximise the nutrient gain from the remnants are all essential parts of carcass handling time.

@@

\subsection*{Food Processing}
Aside from coping with competitors, another vital component of carrion handling time is the ability to maximise the energy gain from the remains while reducing the energetics of doing so. 
At whale carcases, white and blue sharks are known to preferentially feed on the blubber layer \citep{long1996sharks}. 
Blubber is an energy rich portion of the carcass that can allow a shark to survive for 1.5 months on 30 kg of the material \citep{carey1982temperature}. 
On land many scavengers utilize late-stage carcass material that is less subject to decomposition and may be unavailable to other competitors, for example bone.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{timeline_figure/timeLine.pdf}
\caption{The diversity of scavengers through time across air, land and sea. Each species has either direct evidence for it being a scavenger or would be positioned high up on our scavenging scale. The ticks on the x axis represent the transitions for the Palaeozoic-Mesozoic (252.17 Mya) and Mesozoic-Cenozoic (66Mya) boundaries}
\label{Timeline}
\end{figure}

@@

\section*{Acknowledgments}
Thanks to Natalie Cooper for highlighting the potential for this review, and to Ara Monadjem and Deirdre McClean for their comments on the manuscript.
